Self Organizing Maps - Great Lakes Pollen Data - Large SOM
========================================================

Libraries and initialization
--------------------------------------------------------
```{r message=FALSE}
set.seed(3980)

require(spdep)
require(kohonen)
require(RColorBrewer)
require(ggplot2)
require(reshape2)
```

Pre-processing
--------------------------------------------------------

First step is to read in the data and convert to percentages. Steps:

- Read in data
- Remove the data from Benfield (until we can figure out what the problem is)
- Convert to percentages
- Sqrt transformation to downweight the over-represented taxa.

```{r}
gl <- read.csv("./Data/GreatLakesAll_500.csv")
sitesf <- as.factor(substr(gl$Sample,0,4))
ages <- gl$YrBP
```

Extract pollen
```{r}
poll <- gl[,3:24]
pollSum <- apply(poll,1,sum)
poll <- poll/pollSum*100
poll.s<- sqrt(poll)
```

Quick boxplot to check values
```{r message=FALSE}
plot.df = melt(poll.s, variable.name = "Taxa", value.name = "Abundance")
x = ggplot(plot.df, aes(x=Taxa, y=Abundance)) + geom_boxplot() 
x = x + ggtitle("All abundance values") + theme(axis.text.x = element_text(angle = 90, hjust = 1))
print(x)
```

Self-organizing map
--------------------------------------------------------
Start here by building the grid. We use a large grid and then re-cluster (k-means) afterwards. 
```{r}
gridX<-31
gridY<-30
grid.som <-somgrid(gridX, gridY, "hexagonal")
```

Build SOM, using standard unsupervised mapping, and plot the codebook vectors
```{r cache=FALSE}
poll.som <- som(as.matrix(poll.s), grid = grid.som, rlen=1000)
plot(poll.som, codeRendering="stars")
```

Visualisation
--------------------------------------------------------

- Fit of SOM during training
```{r}
plot(poll.som, type="changes")
```

- Number of samples per node - note that not all nodes are full
```{r}
plot(poll.som, type="counts")
```

- Distance to neighbors
```{r}
plot(poll.som, type="dist.neighbours")
```

- Shade by mean age
```{r}
myRCBage = function(n, pal="YlGnBu") {
  brewer.pal(n, pal)
}
nnodes = gridX*gridY
nodeAge = rep(NA, nnodes)
for (i in 1:nnodes) {
  nodeID = which(poll.som$unit.classif == i)
  if (length(nodeID) >0) {
    nodeAge[i] = mean(ages[nodeID])
  }
}

plot(poll.som, type="property", property=nodeAge/1000, main="Mean node age in ka BP",
     palette.name=myRCBage, ncolors=9)
#add.cluster.boundaries(poll.som, poll.som.kmean$cluster)
```

- Shade by mean age with interpolation (simple averaging)
```{r}
myRCBage = function(n, pal="YlGnBu") {
  brewer.pal(n, pal)
}
nnodes = gridX*gridY
nodeAge = rep(NA, nnodes)
grid.dist = unit.distances(poll.som$grid)
for (i in 1:nnodes) {
  nodeID = which(poll.som$unit.classif == i)
  if (length(nodeID) >0) {
    nodeAge[i] = mean(ages[nodeID])
  } else {
    nID = which(grid.dist[i,] < 2)
    nodeID = which(poll.som$unit.classif %in% nID)
    nodeAge[i] = mean(ages[nodeID], na.rm=TRUE)
  }
}

plot(poll.som, type="property", property=nodeAge/1000, main="Mean node age in ka BP",
     palette.name=myRCBage, ncolors=9)
#add.cluster.boundaries(poll.som, poll.som.kmean$cluster)
```

## Clusters

Use `skater()` from the **spdep** package to get contiguous clusters. Start by making the neighborhood structure:
```{r}
som.nb = graph2nb(gabrielneigh(grid.som$pts),sym=TRUE)
#plot(som.nb, grid.som$pts)
```

Make minimum spanning tree by converting this to a weight matrix with wieghts based on node dissimilarity:
```{r}
lcosts <- nbcosts(som.nb, poll.som$codes[[1]])
nb.w <- nb2listw(som.nb, lcosts, style="B")
```

Find the minimum spanning tree:
```{r}
nbclus = 6
som.mst <- mstree(nb.w,nbclus-1)
plot(som.mst, grid.som$pts, col=2,       
     cex.lab=.7, cex.circles=0.035, fg="blue")
```

Now find groups with no restriction
```{r}
som.skat <- skater(som.mst[,1:2], poll.som$codes[[1]], ncuts=(nbclus-1))
som.clus = som.skat$groups
table(som.clus)
```

Make a quick skater plot (can drop this):
```{r}
plot(som.skat, grid.som$pts, cex.circles=0.035, cex.lab=.7)
```

And finally plot this back out on the SOM:
```{r}
myRCBcls = function(n, pal="Dark2") {
  brewer.pal(n, pal)
}
plot(poll.som, type="property", property=as.numeric(som.clus), main="Cluster map", 
     palette.name=myRCBcls, ncolors=nbclus, heatkeywidth = .5)
add.cluster.boundaries(poll.som, som.clus)
```

```{r}
plot.df = cbind(melt(poll.som$codes[[1]], varnames = c("Grid","Taxa"),
                     value.name = "Score"), Clus=as.factor(rep(som.clus, 22)))

x = ggplot(plot.df, aes(x=Taxa, y=Score, fill=Clus)) + geom_boxplot() + facet_wrap(~Clus)
x = x + ggtitle("All abundance values") + theme(axis.text.x = element_text(angle = 90, hjust = 1))
print(x)
```

```{r}
plot.df = data.frame(age=nodeAge, cluster=as.factor(som.clus))
x = ggplot(plot.df, aes(x=reorder(cluster, age*-1, FUN=median, na.rm=TRUE), y=age, fill=cluster)) + geom_boxplot()
x = x + theme_bw() + scale_x_discrete("Cluster") + scale_y_continuous("Age BP")
x = x + ggtitle("Cluster occurrence by age")
print(x)
```

Taxon maps
--------------------------------------------------------
How do variables map to the SOM? 
```{r}
myRCBpol = function(n, pal="Greens") {
  brewer.pal(n, pal)
}
taxNames = names(poll)
for (i in 1:length(taxNames)) {
  plot(poll.som, type="property", property=poll.som$codes[[1]][,taxNames[i]], main=taxNames[i], 
       palette.name=myRCBpol, ncolors=9)
  add.cluster.boundaries(poll.som, som.clus)
}
```

Site trajectories
--------------------------------------------------------
```{r}
# allsites = unique(sitesf)
# for (i in 1:length(allsites)) {
#   siteID= which(sitesf==allsites[i])
#   plot(poll.som, type="dist.neighbours", keepMargins=TRUE, main=allsites[i])
#   add.cluster.boundaries(poll.som, som.clus)
#   site.crd <- poll.som$grid$pts[poll.som$unit.classif,][siteID,]
#   lines(site.crd, lwd=2)
#   points(jitter(site.crd))
# }
```
